Spin-orbit-coupling-induced squeezing in the optical lattice clock

### 2018-07-20 soc_squeezing/soc_squeezing_notes.tex
removed incorrect prefactor on Dicke states; use |m> \propto ... instead
modified derivation of S \dot S Hamiltonian added capability to create a fermionic operator by index added method creating projector onto states with an occupied lattice site changed order of arguments for Hamiltonians in FH methods

If the single-particle energy spacings are large compared to the
strength of inter-particle interactions, or equivalently if
$J_0\gtrsim U$, then by the secular approximation we can neglect terms
in $H_{\t{int}}$ which do not conserve the sum of single-particle
energies.  By solving this single-particle energy conservation
condition, one finds that we must have $\set{p,q}\ne\set{r,s}$; that
is, atoms in different clock states can only interact via
\begin{enumerate*}[label=(\roman*)]
\item direct density-density terms, and
\item terms which exchange their momenta.
\end{enumerate*}
The surviving terms in \eqref{eq:H_int_full} are thus
\begin{align}
  H_{\t{int}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{p,\e}^\dag \hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}
    + \hat c_{q,\e}^\dag \hat c_{p,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}^\dag \hat c_{p,\e}
    - \hat c_{q,\e}^\dag \hat c_{q,\g} \hat c_{p,\g}^\dag \hat c_{p,\e}},
\end{align}
where we can use the single-particle pseudo-spin-1/2 Pauli operators
in \eqref{eq:pseudospin} to write
\begin{align}
  H_{\t{int}} = \f{U}{L} \sum_{p,q}
  \sp{\p{\f{\1_q-\sigma_q^z}{2}} \p{\f{\1_p+\sigma_p^z}{2}}
    - \sigma_q^+ \sigma_p^-}.
\end{align}
Up to a global shift in energy, this result is equivalently
\begin{align}
  H_{\t{int}}
  = - \f14 \f{U}{L} \sum_{p,q} \v{\sigma}_q\c\v{\sigma}_p
  = - \f{U}{L} \sum_{p,q} \v s_q \c\v s_p
  = - \f{U}{L} \v S \c \v S,
\end{align}
where
\begin{align}
  \v{\sigma}_p
  \equiv \p{\sigma_p^x,\sigma_p^y,\sigma_p^z},
  &&
  \v s_p \equiv \p{s_p^x,s_P^y,s_p^z},
  &&
  \v S \equiv \f12 \sum_p \v{\sigma}_p.
  \label{eq:H_int}
\end{align}
For a fixed particle number $N$, the collective spin Hamiltonian
$H_{\t{int}}$ has electronic eigenstates $\set{\ket{Sm}}$ with total
(pseudo-)spin $S$ and projection $m$ onto a quantization axis; the
corresponding energies are $\bk{Sm|H_{\t{int}}|S,m}=-\p{U/L}S\p{S+1}$.
In particular, the ground-state $S=N/2$ manifold is spanned by the
Dicke states
\begin{align}
  \ket{m} \equiv \ket{N/2,m}
  = {N \choose N/2+m}^{-1/2} S_+^{N/2+m} \ket{\g}^{\otimes N},
  &&
  S_+ \equiv \sum_n \sigma_n^+,
  \label{eq:dicke_states}
\end{align}
where $n$ indexes an individual atom,
$\sigma_n^+\equiv\hat c_{n,\e}\hat c_{n,\g}$ is an individual
spin-raising operator, and $N/2+m=0,1,\cdots,N$ is the number of
electronic state excitations in the state $\ket{m}$
\cite{swallows2011suppression}.  In words, the Dicke state $\ket{m}$
is a uniform superposition of all states with $N/2+m$ total electronic
excitations.

>>>>

If the single-particle energy spacings are large compared to the
strength of inter-particle interactions, or equivalently if
$J_0\gtrsim U$, then by the secular approximation we can neglect terms
in $H_{\t{int}}$ which do not conserve the sum of single-particle
energies.  By solving this single-particle energy conservation
condition, one finds that we must have $\set{p,q}\ne\set{r,s}$; that
is, atoms in different clock states can only interact via
\begin{enumerate*}[label=(\roman*)]
\item direct density-density terms, and
\item terms which exchange their momenta.
\end{enumerate*}
The surviving terms in \eqref{eq:H_int_full} are thus
\begin{align}
  H_{\t{int}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{p,\e}^\dag \hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}
    + \hat c_{q,\e}^\dag \hat c_{p,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}^\dag \hat c_{p,\e}
    - \hat c_{q,\e}^\dag \hat c_{q,\g} \hat c_{p,\g}^\dag \hat c_{p,\e}},
\end{align}
where we can use the single-particle pseudo-spin-1/2 Pauli operators
in \eqref{eq:pseudospin} to write
\begin{align}
  H_{\t{int}} = \f{U}{L} \sum_{p,q}
  \sp{\p{\f{\1_q-\sigma_q^z}{2}} \p{\f{\1_p+\sigma_p^z}{2}}
    - \sigma_q^+ \sigma_p^-}
  = \f{U}{L} \sum_{p,q}\f14\p{\1_q \1_p - \v\sigma_q\c\v\sigma_p},
  \label{eq:H_int_pauli}
\end{align}
with
\begin{align}
  \v{\sigma}_p \equiv \p{\sigma_p^x,\sigma_p^y,\sigma_p^z}.
\end{align}
For a fixed total particle number $N$, the identity operators in
\eqref{eq:H_int_pauli} contribute only a global shift in energy, which
allows us to more simply write
\begin{align}
  H_{\t{int}} = - \f{U}{L} \v S\c\v S,
  &&
  \v S \equiv \f12 \sum_p \v{\sigma}_p.
\end{align}
This Hamiltonian has electronic eigenstates $\set{\ket{Sm}}$ with
total (pseudo-)spin $S$ and projection $m$ onto a quantization axis;
the corresponding energies are
$\bk{Sm|H_{\t{int}}|S,m}=-\p{U/L}S\p{S+1}$.  In particular, the
ground-state $S=N/2$ manifold is spanned by the Dicke states
\begin{align}
  \ket{m} \equiv \ket{N/2,m} \propto S_+^{N/2+m} \ket{\g}^{\otimes N},
  &&
  S_+ \equiv \sum_n \sigma_n^+,
  \label{eq:dicke_states}
\end{align}
where $n$ indexes an individual atom,
$\sigma_n^+\equiv\hat c_{n,\e}\hat c_{n,\g}$ is an individual
spin-raising operator, and $N/2+m=0,1,\cdots,N$ is the number of
electronic state excitations in the state $\ket{m}$
\cite{swallows2011suppression}.  In words, the Dicke state $\ket{m}$
is a uniform superposition of all states with $N/2+m$ total electronic
excitations.
